\documentclass[12pt]{article}
\usepackage{geometry,amsmath,amssymb, graphicx, natbib, float, enumerate}
\geometry{margin=1in}
\renewcommand{\familydefault}{cmss}
\restylefloat{table}
\restylefloat{figure}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\logit}{\mathrm{logit}}

\begin{document}
\noindent
{\bf BST 140.751 \\ Problem Set 1} \\
\section{Prerequisites}
\begin{enumerate}[1.]
\item Prove DeMorgan's laws. That is, that $(A \cup B ) ^c = A^c \cap B^c$ and $(A \cap B) ^ c = A^c \cup B^c$.
\item Prove that $(A^c)^c = A$.
\item Prove that $(A \cup B) \cap C = (A\cap C) \cup (B\cap C)$.
\item Show that the function $\frac{1}{\beta} \exp(-x / \beta)$ integrates to $1$ over $x > 0$ for $\beta > 0$.
\item Show that the function $\frac{e^{-x}}{\left(1+  e^{-x} \right)^2}$ integrates to $1$ over $-\infty < x < \infty$.
\item Show that the derivative of the function $(1 + e^{-x})^{-1}$ is the function from the previous problem.
\item Plot the functions from the previous 3 problems in R.
\item Argue that $\sum_{x=0}^\infty \frac{e^{-\lambda}\lambda^x}{x!} = 1$ for $\lambda > 0$.
\item Argue that $\sum_{x=0}^\infty p(1 - p)^{x-1} = 1$ 
\end{enumerate}

\section{Basic probability}
\begin{enumerate}[1.]
\item Given only the simple axiomatic foundations, prove the following.
\begin{enumerate}[a.]
\item  $P(\emptyset) = 0$.
\item  $P(E) = 1 - P(E^c)$.
\item  If $A \subset B$ then $P(A) \leq P(B)$.
\item  For any $A$ and $B$, $P(A \cup B) = P(A) + P(B) - P(A \cap B)$.
\item  $P(A \cup B) = 1 - P(A^c \cap B^c)$.
\item   $P(A \cap B^c) = P(A) - P(A \cap B)$. 
\item  $P(\cup_{i=1}^n E_i) \leq \sum_{i=1}^n P(E_i)$ (Bonferroni's inequality).
\item  $P(\cup_{i=1}^n E_i) \geq \max_i P(E_i)$.
\item $P(B\cap A^C) = P(B) - P(B \cap A)$
\end{enumerate}
\item (From Casella and Berger) $P(A) = 1/3$ and $P(B^C) = 1/4$, can $A$ and $B$ be disjoint?
\item Suppose that an influenza epidemic strikes a area. In $17\%$ of
  two parent families at least one of the parents has contracted the
  disease.  In $12\%$ of the families the father has contracted
  influenza while in $6\%$ of the families both the mother and father
  have contracted influenza.
  \begin{enumerate}[a.]
  \item What's the probability that the mother has contracted influenza?
  \item What's the probability that neither the mother nor the father has contracted influenza? 
  \item What's the probability that the mother has contracted influenza but the father has not?
  \item What's the probability that the father has contracted influenza but the mother has not?
  \item What event does the probability one minus the probability that
    both have contracted influenza represent?
  \end{enumerate}
\end{enumerate}

\section{Random variables, univariate densities}
\begin{enumerate}[1.]
\item Suppose $h(x)$ is such that $h(x) > 0$ for
  $x=1,2,\ldots,I$. Argue that $ p(x) = h(x) / \sum_{i=1}^I h(i) $ is
  a valid pmf.
\item Suppose a function $h$ is such that $h>0$ and $c =
  \int_{-\infty}^\infty h(x) dx < \infty$.  Show that $f(x) = h(x) /
  c$ is a valid density.
\item Suppose that, for a randomly drawn subject from a particular
  population, the proportion of a their skin that is covered in
  freckles follows a density that is constant on $[0,1]$. (This is
  called the {\bf uniform density}.) That is, $f(x) = k$ for $0\leq x
  \leq 1$.
  \begin{enumerate}[a.]
  \item Draw this density. What must $k$ be?
  \item Suppose a random variable, $X$, follows a uniform
    distribution. What is the probability that $X$ is between .1 and
    .7? Interpret
    this probability in the context of the problem.
  \item Verify the previous calculation in R. What's the probability that $a < X < b$ for
    generic values $0 < a < b < 1$?
  \item What is the distribution function associated with this density?
  \item What is the median of this density? Interpret the median in the context of the problem.
  \item What is the $95^{th}$ percentile? Interpret this percentile in the context of the problem.
  \item Do you believe that the proportion of freckles on subjects in a
    given population could feasibly follow this distribution? (Why or
    why not.)
  \end{enumerate}
\item  Consider a randomly drawn leaf from a particular tree
  population. The proportion of the leaf that is covered in blight
  follows density that is a right triangle with one vertex at $(0,0)$,
  one at $(0, k)$ and one at $(1,0)$.
  \begin{enumerate}[a.]
  \item What is the value of $k$ that makes this a valid density?
  \item What is the survival function associated with this density? Plug in $.5$ into
    the survival function. What is this quantity in the context of the problem?
  \item What is the probability of drawing a leaf more than 80\% covered in blight?
  \item What is the $95^{th}$ percentile? Interpret this quantile in the context of the problem.
  \item Do you believe that the proportion of blight on leaves could feasibly follow this
    distribution? (Why or why not.)
  \end{enumerate}
\item Let $0 \leq \pi \leq 1$ and $f_1$ and $f_2$ be two continuous
  densities with associated distribution functions $F_1$ and $F_2$ and
  survival functions $S_1$ and $S_2$. Let $g(x) = \pi f_1(x) + (1 - \pi)f_2(x)$.
  \begin{enumerate}[a.]
	\item Show that $g$ is a valid density.
	\item Write the distribution function associated with $g$ in the terms of
          $F_1$ and $F_2$.
        \item Write the survival function associated with $g$ in the
          terms of $S_1$ and $S_2$.
          \item Repeat the previous questions where $\{f_i\}_{i=1}^I$ is a collection of
          densities and $\{\pi_i\}_{i=1}^I$ is a point on the $I$ dimensional simplex
          ($\sum_{i=1}^I \pi_i = 1$ and $0 \leq \pi_i \leq 1$) and $g = \sum_{i=1}^I \pi_i f_i$.
          \end{enumerate}  
\item  Radiologists have created cancer risk summary that, for a given
  population of subjects, follows (a specific instance of) the {\bf logistic} density
  $$
  \frac{e^{-x}}{(1 + e^{-x})^2} ~~~~~~~\mbox{for}~ -\infty < x < \infty.
  $$
  \begin{enumerate}[a.]
  \item Show that this is a valid density.
  \item Calculate the distribution function associated with this density.
  \item What value do you get when you plug $0$ into the distribution function? Interpret this
    result in the context of the problem.
  \item Define the {\em odds} an event with probability $p$
    as $p / (1 - p)$. Prove that the $p^{th}$ quantile from this
    distribution is $\log\{p / (1 - p)\}$; which is the natural log of
    the odds of an event with probability $p$.
  \end{enumerate}
\item Quality control experts estimate that the time (in years) until a specific electronic 
  part from an assembly line fails follows (a specific instance of) the {\bf Pareto} density
$$
\frac{1}{x^2} ~~~~~~~\mbox{for}~ 1 < x < \infty.
$$
\begin{enumerate}[a.]
\item Show that this is a valid density.
\item What is the survival function associated with this density?
  Interpret a value (say 10 years) evaluated in the survival function in the context of the problem.
\item Show that the $p^{th}$ quantile from this density is $1/(1 -p)$. For $p=.8$ interpret this
  value in the context of the problem.
\end{enumerate}
\item  Suppose that a density is of the form $cx^k$ for some constant $k > 1$ and $0 < x < 1$.
  \begin{enumerate}[a.]
  \item Find $c$.
  \item Derive the distribution function for $f$.
  \item Derive a formula for the $p^{th}$ quantile from $f$.
  \item Let $0 \leq a < b \leq 1$. Derive a formula for $P(a < X < b)$.
  \end{enumerate}
\item Suppose that the time in days until hospital discharge for a certain patient population
follows a density $f(x) = c\exp(-x/10)$ for $x > 0$. 
\begin{enumerate}[a.]
	\item What value of $c$ makes this a valid density?
	\item Find the distribution function for this density.
	\item Find the survival function.
	\item Calculate the probability that a person takes longer than 11 days to be discharged.\
	\item What is the median number of days until discharge?
\end{enumerate}
\item The (lower) incomplete gamma function is 
	defined as $\Gamma(k, c) = \int_{0}^c x^{k-1}\exp(-x)dx$.
	By convention $\Gamma(k, \infty)$, the complete gamma function, is written $\Gamma(k)$.
	Consider a density
	$$
	\frac{1}{\Gamma(\alpha)} x^{\alpha - 1} \exp(-x) ~~ \mbox{for} ~~ x > 0
	$$
	where $\alpha$ is a known number.
	\begin{enumerate}[a.] 
	\item Argue that this is a valid density.
	\item Write out the survival function associated with this density using gamma functions
    \item Let $\beta$ be a known number; argue that
    $$
	\frac{1}{\beta^\alpha\Gamma(\alpha)} x^{\alpha - 1} \exp(-x/\beta) ~~ \mbox{for} ~~ x > 0    
    $$ 	
    is a valid density. This is known as the {\bf gamma density}.
	\item Plot the Gamma density for different values of $\alpha$ and $\beta$.
	\end{enumerate}
\item The {\bf Weibull density} is useful in survival analysis. Its form is given by 
$$
\frac{\gamma}{\beta}x^{\gamma - 1}\exp\left(-x^\gamma / \beta\right),
$$
for $x > 0$ and $\gamma$ and $\beta$ are fixed known numbers.
\begin{enumerate}[a.]
	\item Demonstrate that the Weibull density is a valid density.
	\item Calculate the survival function associated with the Weibull density.
	\item Calculate the median of the Weibull density.
	\item Plot the Weibull density for different values of $\gamma$ and $\beta$.
\end{enumerate}
\item The Beta function is given by $B(\alpha, \beta) =  \int_0^1 x^{\alpha-1} (1 - x)^{\beta - 1}$
for $\alpha > 0$ and $\beta > 0$ . It turns out that 
$$
B(\alpha, \beta) = \Gamma(\alpha)\Gamma(\beta)/\Gamma(\alpha + \beta).
$$
The {\bf Beta density} is given by $\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1 - x)^{\beta-1}$ for fixed 
$\alpha > 0$ and $\beta > 0$. This density is useful for 
\begin{enumerate}[a.]
\item Argue that the Beta density is a valid density.
\item Argue that the uniform density is a special case of the beta density.
\item Plot the beta density for different values of $\alpha$ and $\beta$.
\end{enumerate}
\item A famous formula is $e^{\lambda} = \sum_{x=0}^\infty \frac{\lambda^{x}}{x!}$ for any value
	of $\lambda$. Assume that the count of the number of people infected
with a particular disease per year follows a mass function given by 
$$
P(X = x) = \frac{e^{-\lambda} \lambda^x}{x!} ~~ \mbox{for} ~~ x = 0, 1, 2, 3, \ldots
$$
where $\lambda$ is a fixed known number. (This is know as the {\bf Poisson mass function}.)
\begin{enumerate}[a.]
\item Argue that $\sum_{x=0}^\infty P(X = x) = 1$.
\end{enumerate}
\item Consider counting the number of coin flips from an unfair coin with success probability $p$ until a head is obtained, say $X$. The mass
function for this process is given by $P(X = x) = p(1 - p)^{x-1}$ for $x = 1, 2, 3, \ldots$.
This is called the {\bf geometric mass function}.
\begin{enumerate}[a.]
\item Argue mathematically that this is a valid probability mass function. Hint, the geometric series is given by $\frac{1}{1-r} = \sum_{k=0}^\infty r^k$ for $|r| < 1$. 
\item Calculate the survival distribution $P(X > x)$ for the geometric distribution for
integer values of $x$.
\end{enumerate}
\item Let $U$ be a uniform $(0,1)$ density. Calculate the distribution function and density of $U^P$ where $p$ is a power. What is the name of this density?
\item Let $X$ be an exponential $1$ random variable. Calculate the distribution and density for $\log(X)$.

\end{enumerate}



\section{Expected values and variances}
\begin{enumerate}[1.]
\item  Using the rules of expectations prove that $\Var(X) = E[X^2] - E[X]^2$
  where $\Var(X) = E[(X - \mu)^2]$.
\item Let $g(x) = \pi f_1(x) + (1 - \pi)f_2(x)$ where $f_1$ and $f_2$
  are densities with associated means and variances $\mu_1$,
  $\sigma^2_1$ and $\mu_2$, $\sigma^2_2$, respectively. You showed
  already that $g$ is a valid density. What is it's associated mean
  and variance?
\item  Suppose that a density is of the form $(k + 1)x^k$ for some constant $k > 1$ and $0 < x < 1$.
  \begin{enumerate}[a.]
  \item What is the mean of this distribution?
  \item What is the variance?
  \end{enumerate}
\item Suppose that the time in days until hospital discharge for a certain patient population
follows a density $f(x) = \frac{1}{10}\exp(-x/10)$ for $x > 0$. 
\begin{enumerate}[a.]
\item Find the mean and variance of this distribution?
\item The general form of this density (the exponential density) is $f(x) = \frac{1}{\beta}\exp(-x/\beta)$ for $x > 0$ for a fixed value of $\beta$. Calculate the mean and variance of this density.
\end{enumerate}
\item The Gamma density is given by     
	$$
	\frac{1}{\beta^\alpha\Gamma(\alpha)} x^{\alpha - 1} \exp(-x/\beta) ~~ \mbox{for} ~~ x > 0    
    $$ 	
for fixed values of $\alpha$ and $\beta$.
\begin{enumerate}[a.]
\item Derive the mean and variance of the gamma density. You can assume the fact (proved in HW 1) that the density integrates to 1 for any $\alpha > 0$ and $\beta > 0$.
\item The Chi-squared density is the special case of the Gamma density where $\beta = 2$ and 
$\alpha = p / 2$ for some fixed value of $p$ (called the ``degrees of freedon''). Calculate the
mean and variance of the Chi-squared density.
\end{enumerate}
\item The Beta density is given by 
$$
\frac{1}{B(\alpha, \beta)} x^{\alpha - 1}(1 - x)^{\beta - 1} ~~\mbox{for}~~ 0 < x< 1
$$
and
$
B(\alpha, \beta) = \Gamma(\alpha)\Gamma(\beta)/\Gamma(\alpha + \beta).
$
\begin{enumerate}[a.]
\item Derive the mean of the beta density. Note the following is useful for simplifying results: $\Gamma(c + 1) = c\Gamma(c)$ for $c > 0$.
\item Derive the variance of the beta density.
\end{enumerate}
\item The Poisson mass function is given by
$$
P(X = x) = \frac{e^{-\lambda} \lambda^x}{x!} ~~ \mbox{for} ~~ x = 0, 1, 2, 3, \ldots
$$
\begin{enumerate}[a.]
\item Derive the mean of this mass function.
\item Derive the variance of this mass function. Hint, consider $E[X(X - 1)]$.
\end{enumerate}
\item Suppose that, for a randomly drawn subject from a particular
  population, the proportion of a their skin that is covered in
  freckles follows a uniform density (constant between $0$ and $1$).  
  \begin{enumerate}[a.]
  \item What is the expected percentage of a (randomly selected) person's body that is covered in freckles? (Show your work.)
  \item What is the variance? (Show your work.) 
  \end{enumerate}
\item  You have an MP3 player with a total of 4 songs stored on it. Suppose that songs are played
  randomly {\em with replacement}. Let $X$ be the number of songs played
  until you hear a repeated song. 
  \begin{enumerate}[a.]
  \item What values can $X$ take, and with what probabilities?
  \item What is the expected value for $X$?
  \item What is the variance for $X$?
  \end{enumerate}
\item Argue that any density that is symmetric about a point $\mu$, 
  whose first moment exists, has mean $\mu$. 
\item Give a proper density that has no mean. Give a proper density
  that has a mean but no variance.
\item (Adapted from Casella and Berger) 
  Let $x_0$ be a point. Argue that $g(x) = f(x) / F(X_0)$ for $x < x_0$ is
  a valid density. (This is the conditional density given that $x < x_0$.)
  What is the CDF associated with $g$? 
\item Let $X$ be a random variable with mean $\mu$ and variance $\sigma^2$. Show
that $(X- \mu) / \sigma$ has mean $0$ and variance $1$.

\item You are playing a game with a friend where you flip a coin and if it comes up heads you
  give him a dollar and if it comes up tails she gives you a dollar. You play the game ten times.
  \begin{enumerate}[a.]
  \item What is the expected total earnings for you? (Show your work; state your assumptions.)
  \item What is the variance of your total earnings? (Show your work; state your assumptions.)
  \item Suppose that the die is biased and you have a $.4$ chance of winning for each flip.
    repeat the calculations in parts $a$ and $b$
  \end{enumerate}
 \item Note that the code 
\begin{verbatim}
temp <- matrix(sample(1 : 6, 1000 * 10, replace = TRUE), 1000)
xBar <- apply(temp, 1, mean)
\end{verbatim}
    In R produces $1,000$ averages of $10$ die rolls. That is, it's
    like taking ten dice, rolling them, averaging the results and
    repeating this $1,000$ times.
    \begin{enumerate}[a.]
    \item Do this in R. Plot histograms of the averages.
    \item Take the mean of \texttt{xBar}. What should this value be close to? (Explain your reasoning.)
    \item Take the standard deviation of \texttt{xBar}. What should this value be close to? (Explain your reasoning.)
    \end{enumerate}
  \item Note that the code
\begin{verbatim}
xBar <- apply(matrix(runif(1000 * 10), 1000), 1, mean)
\end{verbatim}
    produces $1,000$ averages of $10$ uniforms.
    \begin{enumerate}[a.]
    \item Do this in R. Plot histograms of the averages.
    \item Take the mean of \texttt{xBar}. What should this value be close to? (Explain your reasoning.)
    \item Take the standard deviation of \texttt{xBar}. What should this value be close to? (Explain your reasoning.)
    \end{enumerate}
\item You flip a coin with a probability $p$ of heads. How many times do you have to flip it on average
  until you get a head? What's the variance of the number of flips to get a head?
\item Let $X$ be a random variable such that $0 \leq X \leq 1$. What's the largest that the variance of
$X$ can be? What distribution for $X$ obtains this variance?
\end{enumerate}


\section{Random vectors, independence, conditional probabilities}
\begin{enumerate}[1.]
\item When at the free-throw line for two shots, a basketball player
  makes at least one free throw 90\% of the time. 80\% of the time,
  the player makes the first shot, while 70\% of the time she makes
  both shots.
\begin{enumerate}[a.]
\item Does it appear that the player's second shot success is
  independent of the first?
\item What is the conditional probability that the player makes the
  second shot given that she made the first? What would it be if she
  missed the first?
\end{enumerate}
\item Assume that an act of intercourse between an HIV infected person
  and a non-infected person results in a $1/500$ probability of
  spreading the infection. How many acts of intercourse would an
  uninfected person have to have with an infected persons to have a
  50\% probability of obtaining an infection?  State the assumptions
  of your calculations.
\item You meet a person at the bus stop and strike up a
  conversation. In the conversation, it is revealed that the person is
  a parent of two children and that one of the two children is a
  girl. However, you do not know the gender of the other child, nor
  whether the daughter she mentioned is the older or younger sibling.
\begin{enumerate}[a.]
\item What is the probability that the other sibling is a girl? What
  assumptions are you making to perform this calculation?
\item Later in the conversation, it becomes apparent that she was
  discussing the older sibling.  Does this change your probability
  that the other sibling is a girl?
\end{enumerate}
\item A particularly sadistic warden has three prisoners, A, B and
  C. He tells prisoner C that the sentences are such that two
  prisoners will be executed and let one free, though he will not say
  who has what sentence. Prisoner C convinces the warden to tell him
  the identity of one of the prisoners to be executed. The warden has
  the following strategy, which prisoner C is aware of.  If C is
  sentenced to be let free, the warden flips a coin to pick between A
  and B and tells prisoner C that person's sentence. If C is sentenced
  to be executed he gives the identity of whichever of A or B is also
  sentenced to be executed.
\begin{enumerate}[a.]
\item Does this new information about one of the other prisoners give
  prisoner C any more information about his sentence?
\item The warder offers to let prisoner C switch sentences with the
  other prisoner whose sentence he has not identified. Should he
  switch?
\end{enumerate}
\item Derive the mean and variance for the difference in means
  between a collection of data $\{X_{1i}\}_{i=1}^I$ and $\{X_{2i}\}_{i=1^I}$.
  Where $(X_{i1}, X_{i2})$ are iid pairs of data (yet possibly dependent).
  Define any notation that you use. How does the variance change
  when you assume that the observations within a pair are independent.
  \item  Quality control experts estimate that the time (in years) until a specific electronic 
    part from an assembly line fails follows (a specific instance of) the {\bf Pareto} density
    $$
    \frac{3}{x^4} ~~~~~~~\mbox{for}~ 1 < x < \infty.
    $$
    \begin{enumerate}[a.]
    \item What is the average failure time for components from this density? (Show your work.)
    \item What is the variance? (Show your work.)
	\item The general form of the Pareto density is given by $\frac{\beta \alpha^\beta}{x^{\beta + 1}}$
for $0 < \alpha < x$ and $\beta > 0$ (for fixed $\alpha$ and $\beta$). Calculate the mean and variance of the general Pareto density.
    \end{enumerate}
\end{enumerate}

\section{Conditional probabilities, Bayes rule}
\begin{enumerate}[1.]
\item  A web site (www.medicine.ox.ac.uk/bandolier/band64/b64-7.html) for home pregnancy tests cites the following:
\begin{quote}
  When the subjects using the test were women who collected and tested
  their own samples, the overall sensitivity was 75\%. Specificity was
  also low, in the range 52\% to 75\%.
\end{quote}
\begin{enumerate}[a.]
\item Interpret a positive and negative test result using diagnostic
  likelihood ratios using both extremes of the specificity.
\item A woman taking a home pregnancy test has a positive test. Draw a
  graph of the positive predictive value by the prior probability
  (prevalence) that the woman is pregnant. Assume the specificity is 63.5\%
\item Repeat the previous question for a negative test and the negative
  predictive value.
\end{enumerate}

\section{Standard distributions}
\begin{enumerate}[1.]
\item (Adapted from Rosner page 135) Suppose that the diasolic blood
  pressures of $35-44$ year old men are normally distributed with mean
  $80$ ($mm$ $Hg$) and variance $144$. For the same population, the
  systolic blood pressures are also normally distributed and have a
  mean of $120$ and variance $121$.
  \begin{enumerate}[a.]
  \item What is the probability that a randomly selected person from
    this population has a DBP less than 90?
  \item What DBP represents the $90^{th}$, $95^{th}$ and $97.5^{th}$
    percentiles of this distribution?
  \item What's the probability of a random person from this population
    having a SBP $1$, $2$ or $3$ standard deviations above $120$? 
    What's the corresponding probabilities for having DBPs $1$, $2$ or
    $3$ standard deviations above $80$?
  \item Suppose that $10$ people are sampled from this population.  What's
    the probability that 50\% (5) of them have a SBP larger than $140$?
  \item Suppose that $1,000$ people are sampled from this population.  What's
    the probability that 50\% (500) of them have a SBP larger than $140$?
  \item If a person's SBP and DBP are independent, what's the
    probability that a person has a SBP larger than $140$ and a DBP
    greater than $90$? Is independence a good assumption?
  \item Suppose that an average of $200$ people are drawn from 
    this population. What's the probability that this average
    is smaller than $81.3$?
  \end{enumerate}  
\item  Suppose that IQs in a particular population are normally
  distributed with a mean of $110$ and a standard deviation of $10$.
  \begin{enumerate}[a.]
  \item What's the probability that a randomly selected person from
    this population has an IQ between $95$ and $115$?
  \item What's the $65^{th}$ percentile from this distribution?
  \item Suppose that $5$ people are sampled from this distribution. 
    What's the probability $4$ (80\%) or more have IQs above $130$?
  \item Suppose that $500$ people are sampled from this distribution. 
    What's the probability $400$ (80\%) or more have IQs above $130$?
  \item Consider the average of $100$ people drawn from this
    distribution. What's the probability that this mean is larger
    than $112.5$?
  \end{enumerate}
\item  Suppose that $400$ observations are drawn at random
  from a distribution with mean $0$ and standard deviation $40$.
  \begin{enumerate}[a.]
  \item What's the approximate probability of getting a sample mean larger
    than $3.5$?
  \item Was normality of the underlying distribution required for this
    calculation?
  \end{enumerate}
\end{enumerate}

\section{Limit theorems and sampling distributions}
\begin{enumerate}[1.]
\item Recall that R's function \texttt{runif} generates (by default)
  random uniform variables that have means $1/2$ and variance $1/12$.
  \begin{enumerate}[a.]
  \item Sample $1,000$ observations from this distribution. Take the
    sample mean and sample variance. What numbers should these
    estimate and why?
  \item Retain the same $1,000$ observations from part a. Plot the
    sequential sample means by observation number. Hint. If $x$ is a
    vector containing the simulated uniforms, then the code \texttt{y
      <- cumsum(x) / (1 : length(x))} will create a vector of the
    sequential sample means. Explain the resulting plot.
  \item Plot a histogram of the $1,000$ numbers. Does it look like a
    uniform density?
  \item Now sample $1,000$ {\em sample means} from this distribution,
    each comprised of $100$ observations. What numbers should the
    average and variance of these $1,000$ numbers be equal to and why?
    Hint. The command 
    \begin{quote}
      \texttt{x <- matrix(runif(1000 * 100), nrow = 1000)}      
    \end{quote}
    creates a matrix of size $1,000\times 100$ filled with random
    uniforms. The command \texttt{y<-apply(x,1,mean)} takes the sample
    mean of each row.
  \item Plot a histogram of the $1,000$ sample means appropriately
    normalized. What does it look like and why?
  \item Now sample $1,000$ {\em sample variances} from this distribution,
    each comprised of $100$ observations. Take the average of these
    $1,000$ variances. What property does this illustrate and why?
  \end{enumerate}
\item Note that R's function \texttt{rexp} generates random
  exponential variables. The exponential distribution with rate $1$
  (the default) has a theoretical mean of $1$ and variance of $1$.
  \begin{enumerate}[a.]
  \item Sample $1,000$ observations from this distribution. Take the
    sample mean and sample variance. What numbers should these
    estimate and why?
  \item Retain the same $1,000$ observations from part a. Plot the
    sequential sample means by observation number. Explain the
    resulting plot.
  \item Plot a histogram of the $1,000$ numbers. Does it look like a
    exponential density?
  \item Now sample $1,000$ {\em sample means} from this distribution,
    each comprised of $100$ observations. What numbers should the
    average and variance of these $1,000$ numbers be equal to and why?
 \item Plot a histogram of the $1,000$ sample means appropriately
    normalized. What does it look like and why?
  \item Now sample $1,000$ {\em sample variances} from this distribution,
    each comprised of $100$ observations. Take the average of these
    $1,000$ variances. What property does this illustrate and why?
  \end{enumerate}
\item  Consider the distribution of a fair coin flip (i.e. a random variable
  that takes the values $0$ and $1$ with probability $1/2$ each.)
  \begin{enumerate}[a.]
  \item Sample $1,000$ observations from this distribution. Take the
    sample mean and sample variance. What numbers should these
    estimate and why?
  \item Retain the same $1,000$ observations from part a. Plot the
    sequential sample means by observation number. Explain the
    resulting plot.
  \item Plot a histogram of the $1,000$ numbers. Does it look like it places
    equal probability on $0$ and $1$?
  \item Now sample $1,000$ {\em sample means} from this distribution,
    each comprised of $100$ observations. What numbers should the
    average and variance of these $1,000$ numbers be equal to and why?
 \item Plot a histogram of the $1,000$ sample means appropriately
    normalized. What does it look like and why?
  \item Now sample $1,000$ {\em sample variances} from this distribution,
    each comprised of $100$ observations. Take the average of these
    $1,000$ variances. What property does this illustrate and why?
  \end{enumerate}  
\item Consider a density for the proportion of a person's body that is
  covered in freckles, $X$, given by $f(x) = cx$ for $0 \leq x \leq 1$
  and some constant $c$.
  \begin{enumerate}[a.]
  \item What value of $c$ makes this function a valid density? 
  \item What is the mean and variance of this density?
  \item You simulated $100,000$ sample means, each comprised of $100$
    draws from this density. You then took the variance of those
    $100,000$ numbers. Approximately what number did you obtain?
    (Explain.)
  \end{enumerate}
\item Suppose that DBPs drawn from a certain population are normally
    distributed with a mean of $90$ $mmHg$ and standard deviation of $5$
    $mmHg$. Suppose that $1,000$ people are drawn from this population.
  \begin{enumerate}[a.]
  \item If you had to guess the number of people people in having DBPs
    less than $80$ $mmHg$ what would you guess?
  \item  You draw
    $25$ people from this population. What's the probability tha the
    sample average is larger than $92$ $mmHg$? 
  \item  You select $5$
    people from this population. What's the probability that $4$ or
    more of them have a DBP larger than $100$ $mmHg$? 
  \end{enumerate}
\item You need to calculate the probability that a {\em standard normal} is 
  larger than $2.20$, but have nothing available other than a regular coin.
  Describe how you could estimate this probability using only your coin. (Do not
  actually carry out the experiment, just describe how you would do it.)
\item Let $X_1$, $X_2$ be independent, identically distributed coin
  flips (taking values $0$ = failure or $1$ = success) having success
  probability $\pi$. Give and interpret the likelihood ratio comparing
  the hypothesis that $\pi = .5$ (the coin is fair) versus $\pi = 1$ (the
  coin always gives successes) when both coin flips result in
  successes.
\item The density for the population of increases in wages for
assistant professors being promoted to associates (1 = no increase, 2 = salary has
doubled) is uniform on the range from 1 to 2.
\begin{enumerate}[a.]
\item What's the mean and variance of this density?
\item Suppose that the sample variance of $10$ observations from this
  density was sampled say $10,000$ times. What number would we expect
  the average value from these $10,000$ variances to be near? (Explain
  your answer briefly.)
\end{enumerate}
\item Suppose that the US intelligence quotients (IQs) are normally
  distributed with mean $100$ and standard deviation $16$.
  \begin{enumerate}[a.]
  \item  What IQ
    score represents the $5^{th}$ percentile? (Explain your calculation.)
  \item Consider the previous question. Note that $116$ is the
    $84^{th}$ percentile from this distribution. Suppose now that
    $1,000$ subjects are drawn at random from this population. Use the
    central limit theorem to write the probability that less than
    $82\%$ of the sample has an IQ below $116$ as a standard normal
    probability. Note, you do not need to solve for the final number. (Show your work.) 
  \item Consider the previous two questions. Suppose now that a sample
    of $100$ subjects are drawn from a {\em new} population and that
    $60$ of the sampled subjects had an IQs below $116$. Give a $95\%$
    confidence interval estimate of the true probability of drawing a
    subject from this population with an IQ below $116$. Does this proportion
    appear to be different than the $84\%$ for the population from questions 1 and 2?
  \end{enumerate}
\item Let $X$ be binomial with success probability $p_1$ and $n_1$
    trials and $Y$ be an independent binomial with success probability
    $p_2$ and $n_2$ trials.  Let $\hat p_1 = X / n_1$ and $\hat p_2 =
    Y / n_2$ be the associated sample proportions. What would be an
    estimate for the standard error for $\hat p_1 - \hat p_2$? To have
    consistent notation with the next problem, label
    this value $\hat{SE}_{\hat p_1 - \hat p_2}$.
\item You are in desperate need to simulate standard normal random
  variables yet do not have a computer available. You do, however, have
  ten standard six sided dice. Knowing that the mean of a single die roll
  is $3.5$ and the standard deviation is $1.71$, describe how you 
  could use the dice to approximately simulate standard normal random variables. (Be precise.)
\item \item Consider three sample variances, $S_1^2$, $S_2^2$ and
  $S_3^2$. Suppose that the sample variances are comprised of $n_1$,
  $n_2$ and $n_3$ iid draws from normal populations $N(\mu_1,
  \sigma^2)$, $N(\mu_2, \sigma^2)$ and $N(\mu_3, \sigma^2)$, respectively. Argue
  that $$\frac{(n_1 - 1)S_1^2 + (n_2 - 1)S_2^2 + (n_3 - 1) S_3^2}{n_1
    + n_2 + n_3 - 3}$$ is an unbiased estimate of $\sigma^2$.
\item You need to calculate the probability that a normally
  distributed random variable is less than $1.25$ standard deviations
  below the mean. However, you only have an oddly shaped coin with a
  known probability of heads of $.6$. Describe how you could estimate
  this probability using this coin. (Do not actually carry out the
  experiment, just describe how you would do it.)
\item The next three questions (A., B., C.) deal with the following
  setting. Forced expiratory volume, $FEV_1$, is a measure of lung
  function that is often expressed as a proportion of lung capacity
  called forced vital capacity, FVC.  Suppose that the population
  distribution of $FEV_1/FVC$ of asthmatics adults in the US has mean
  of $.55$ and standard deviation of $.10$.
\begin{enumerate}[a.]
\item Suppose a random sample of $100$ people are drawn from this
  population. What is the probability that their average $FEV_1/FVC$ is
  larger than $.565$?
\item Suppose the population of non-asthmatics adults in the US have
  a mean $FEV_1/FVC$ of $.8$ and a standard deviation of $.05$.
  You sample $100$ people from the asthmatic population and
  $100$ people from the non-asthmatic population and take the
  difference in sample means. You repeat this process $10,000$
  times to obtain $10,000$ differences in sample means. What
  would you guess the mean and standard deviation of these
  $10,000$ numbers would be?
\item Moderate or severe lung dysfunction is defined as $FEV_1/FVC
  \leq .40$. A colleague tells you that $60\%$ of asthmatics
  in the US have moderate or severe lung dysfunction. To verify this,
  you take a random sample of $5$ subjects, only one of which has
  moderate or severe lung dysfunction. What is the probability of
  obtaining only one or fewer if your friend's assertion is
  correct? What does your result suggest about their
  assertion? 
\end{enumerate}

\end{enumerate}

\section{Likelihood}
\item Imagine that a person, say his name is Flip, has an oddly
  deformed coin and tries the following experiment.  Flip flips his
  coin $10$ times, $7$ of which are heads. You think maybe Flip's coin
  is biased towards having a greater probability of yielding a head than 50\%.
  \begin{enumerate}[a.]
  \item What is the maximum likelihood estimate of $p$, the true probability of
    heads associated with this coin? 
  \item Plot the likelihood associated with this experiment. Renormalize the
    likelihood so that its maximum is one. Does the likelihood suggest that
    the coin is fair?
  \item Whats the probability of seeing $7$ or more heads out of ten
    coin flips if the coin was fair? Does this probability suggest
    that the coin is fair? Note this number is called a P-value.
  \item Suppose that Flip told you that he did not fix the number of
    trials at $10$. Instead, he told you that he had flipped the coin
    until he obtained $3$ tails and it happened to take $10$ trials to
    do so. Therefore, the number $10$ was random while the number
    three $3$ fixed.  The probability mass function for the number of
    trials, say $y$, to obtain $3$ tails (called the negative binomial
    distribution) is
    $$
    \left(\begin{array}{c}y-1 \\ 2\end{array}\right)(1-p)^3p^{y - 3}
    $$
    for $y=3,4,5,6,\ldots$. What is the maximum likelihood estimate of
    $p$ now that we've changed the underlying mass function?
  \item Plot the likelihood under this new mass function. Renormalize
    the likelihood so that its maximum is one. Does the likelihood
    suggest that the coin is fair?
  \item Calculate the probability of requiring $10$ or more flips to
    obtain $3$ tails if the coin was fair. (Notice that this is the
    same as the probability of obtaining $7$ or more heads to obtain
    $3$ tails.) This is the Pvalue under
    the new mass function.
    \ \\ \ \\
    (Aside) This problem highlights a distinction between the
    likelihood and the P-value. The likelihood and the MLE are the
    same regardless of the experiment. That is to say, the likelihood
    only seems to care that you saw $10$ coin flips, $7$ of which were
    heads. Flip's intention about when he stopped flipping the coin,
    either at $10$ fixed trials or until he obtained $3$ tails, are
    irrelevant as far as the likelihood is concerned. The P-value, in
    comparison, does depend on Flip's intentions.
  \end{enumerate}
\item  Suppose a researcher is studying the number of sexual
  acts with an infected person until an uninfected person contracts an
  sexually transmitted disease. She assumes that each encounter is an
  independent Bernoulli trial with probability $p$ that the subject
  becomes infected. This leads to the so-called geometric distribution
  $P(\mbox{Person is infected on contact } x) = p(1 - p)^{x-1}$ for
  $x=1,\ldots$.
  \begin{enumerate}[a.]
  \item Suppose that one subject's number of encounters until
    infection is recorded, say $x$. Symbolically derive the ML
    estimate of $p$.
  \item Suppose that the subjects value was $2$. Plot and interpret
    the likelihood for $p$.
  \item Suppose that is often assumed that the probability of
    transmission, $p$, is $.01$. The researcher thinks that it is
    perhaps strange to have a subject get infected after only $2$
    encounters if the probability of transmission is really on
    $1\%$. According to the geometric mass function, what is the
    probability of a person getting infected in $2$ or fewer
    encounters if $p$ truly is $.01$?
  \item Suppose that she follows $n$ subjects and records the number
    of sexual encounters until infection (assume all subjects became
    infected) $x_1,\ldots,x_n$. Symbolically derive the ML
    estimate of $p$.
  \item Suppose that she records values $x_1 = 3$, $x_2 = 5$, $x_3 = 2$.
    Plot and interpret the likelihood for $p$.
  \end{enumerate}  
\item  In a study of aquaporins $6$ frog eggs received a protein
  treatment. If the treatment of the protein is effective, the frog
  eggs would implode. The experiment resulted in $5$ frog eggs
  imploding.  Historically, ten percent of eggs implode without the
  treatment. Assuming that the results for each egg are independent
  and identically distributed:
  \begin{enumerate}[a.]
  \item What's the probability of getting $5$ or more eggs imploding
    in this experiment if the true probability of implosion is $10\%$?
    Interpret this number.
  \item What is the maximum likelihood estimate for the probability of
    implosion?
  \item Plot and interpret the likelihood for the probability of implosion.
  \end{enumerate}
\item Consider a sample of $n$ iid draws from an exponential density
$$
\frac{1}{\beta}\exp(-x / \beta) ~~\mbox{for}~~ \beta > 0.
$$
\begin{enumerate}[A.]
\item Derive the maximum likelihood estimate for $\beta$.
\item Suppose that in your experiment, you obtained five
observations 
\begin{verbatim}
 1.590 0.109 0.155 0.281 0.453
\end{verbatim}
plot the likelihood for $\beta$. Put in reference lines at 1/8 and 1/16.
\end{enumerate}
\item Often infection rates per time at risk are modelled as Poisson
random variables. Let $X$ be the number of infections and let $t$ be the 
person days at risk. Consider the Poisson mass function
$
(t\lambda)^x \exp(-t\lambda) / x!
$
The parameter $\lambda$ is called the population incident rate.
\begin{enumerate}[A.]
\item Derive the ML estimate for $\lambda$.
\item Suppose that 5 infections are recorded per 1000 person-days at risk. Plot the likelihood.
\item Suppose that five independent hospitals are monitored and that the infection rate ($\lambda$) is assumed to be the same at all five. Let $X_i$, $t_i$ be the count of the number of infections
and person days at risk for hospital $i$. Derive the ML estimate of $\lambda$.
\end{enumerate}
\item Consider $n$ iid draws from a gamma density where $\alpha$ is known
$$
\frac{1}{\Gamma(\alpha)\beta^\alpha} x^{\alpha -1} \exp(-x/\beta) ~~~\mbox{for}~~~\beta > 0, x>0, \alpha > 0.
$$
\begin{enumerate}[A.]
\item Derive the ML estimate of $\beta$.
\item Suppose that $n=5$ observations were obtained: 0.015, 0.962, 0.613, 0.061, 0.617. Draw
	a likelihood plot for $\beta$ (still assume that $\alpha = 1$).
\end{enumerate}
\item Let $Y_1,\ldots, Y_N$ be iid random variabels from a Lognormal distribution with parameters
	$\mu$ and $\sigma^2$. Note $Y \sim \mbox{Lognormal}(\mu,\sigma^2)$ if and only if
	$\log Y \sim N(\mu,\sigma^2)$. The log-normal density is given by
	$$
	(2\pi \sigma^2)^{-1/2} \exp[-\{\log(y) - \mu\}^2 / 2\sigma^2] / y ~~\mbox{for}~~ y > 0
	$$
	\begin{enumerate}[A.]
	\item Show that the ML estimate of $\mu$ is $\hat \mu = \frac{1}{N} \sum_{i=1}^N \log(Y_i)$. (The
	mean of the log of the observations. This is called the ``geometric mean''.)
	\item Show that the ML estimate of $\sigma^2$ is then the biased variance estimate based
	on the log observation
	$$
	\frac{1}{N}\sum_{i=1}^N (\log(y_i) - \hat \mu)^2
	$$
	\end{enumerate}
\end{enumerate}

\section{Programming}
\begin{enumerate}[1.]
\item Simulate $1,000$ random uniform[0,1] variables
  variables and do the following:
  \begin{enumerate}[a.]
  \item Calculate their mean, variance and standard deviation.
  \item Plot a histogram.
  \end{enumerate}
\item Take your $1,000$ simulated uniform random variables, say the
  vector $U$, and create the vector $Y = 2 + 4 * U + E$ where $E$ is
  $1,000$ simulated standard normal random variables.
  \begin{enumerate}[a.]
  \item Plot $Y$ versus $U$
  \item Use $lm(Y ~ U)$ to find the line that best fits the data.
  \item Overlay the line onto your plot.
  \end{enumerate}
\item Write a function in R that takes in a matrix and normalizes it so that
  every row and column has a mean of 0 and a variance of 1. Note, to normalize
  a vector, subtract its mean from every element then divide every resulting difference
  by the standard deviation. Test your function by simulating matrices of random uniforms.
\item Use R to create a graph of an Archimedian spiral.
\item Graph Tupper's self referential formula in R.
\item Create a data frame in R using the $Y$ and $U$ from above. Add a third variable
  that is $2Y$. Change the names of the variables. 
\item Write a function in R that takes in a character vector of first and last names
  in the format {\bf firstname.lastname} and returns a character vector of the form
{\bf lastname.firstname}.
\item Create three random $2\times 2$ matrices consisting of uniform variables. 
  Create a (three element) list of these matrices. Use an lapply statement to create
  a new list consisting of the matrices squared.
\end{enumerate}
\end{document}
